
# Enter Chapter title here {#ch7}

## Summary of Lecture 7 – Fixed Income Valuation and Term Structure

### Learning Objectives
- Understand bond pricing fundamentals and the relationship between bond prices, yield to maturity, and coupon rates.
- Valuation and characteristics of zero-coupon bonds and building the yield curve.
- Valuing coupon-bearing bonds using present value of future cash flows.
- Factors affecting yield curve shape and deriving forward rates from the yield curve.

### Bond Pricing Models

#### Bond Pricing Model I
The price of a bond is the present value of expected cash flows, including coupon and principal. The discount rate is the yield to maturity ($YTM$).

$$
\begin{align}
P_b &= \sum_{t=1}^{T} \frac{C_t}{(1 + r)^t} + \frac{F}{(1 + r)^T}
\end{align}
$$



$$
\text{Annualities Factor} = \frac{1}{r} \left( 1 - \frac{1}{(1 + r)^T} \right)\
$$

$$
P_B = C_t \times \text{Annuity Factor} + \text{Par} \times \text{PV Factor}\frac{C_t}{r} \left[ 1 - \frac{1}{(1 + r)^T} \right] + \frac{\text{Par}}{(1 + r)^T}
$$

Where:

- $P_b$ = price of the bond
- $C_t$ = coupon payment
- $F$ = face value of the bond
- $T$ = number of periods
- $r$ is the appropriate discount rate, required return by bondholders, or yield to maturity 

```{r}

calculate_present_value <- function(C_t, r, T, Par) {
  annuity_factor <- (1 - 1 / (1 + r)^T) / r
  pv_factor <- 1 / (1 + r)^T
  P_B <- C_t * annuity_factor + Par * pv_factor
  return(P_B)
}

# Example usage
C_t <- 1000
r <- 0.05
T <- 10
Par <- 10000

P_B <- calculate_present_value(C_t, r, T, Par)
P_B
```
```{r}
# Define the bond characteristics
par_value <- 10000000
coupon_rate <- 0.08
coupon <- par_value * coupon_rate
market_price <- 10302284.43
n_years <- 3

# Define the function for calculating bond price given a yield (ytm)
bond_price <- function(ytm) {
  price <- coupon / (1 + ytm) + 
           coupon / (1 + ytm)^2 + 
           (coupon + par_value) / (1 + ytm)^3
  return(price)
}

# Define the function to find the root (market_price - bond_price)
find_ytm <- function(ytm) {
  bond_price(ytm) - market_price
}

# Use uniroot to solve for ytm
result <- uniroot(find_ytm, lower = 0, upper = 1)
ytm <- result$root
ytm
```
#### Bond Pricing Model II Zero-coupon bonds

Zero-coupon bonds involve calculating the present value of the par value, discounted by the zero-coupon rate.

$$
\begin{align}
P_0 &= \frac{F}{(1 + z_{0T})^T}
\end{align}
$$

Where:

$F$ace value or principal
$P_0$: current (time 0) bond price
$Z_{0T}$: zero-coupon rate (p.a.), or “zero coupon rate”, from time 0 to time T


```{r}
# Define values
par_value <- 1000000  # Par value
P0 <- 650000          # Price
T <- 5                # Term to maturity (years)

# Calculate implied zero-coupon rate
zero_rate <- (par_value / P0)^(1 / T) - 1
cat("The implied 5-year zero rate is:", round(zero_rate * 100, 2), "% p.a.\n")

# Comparative statics
new_par_value <- 1200000
new_P0 <- 700000
new_T <- 6

# New implied zero rates
new_zero_rate_par <- (new_par_value / P0)^(1 / T) - 1
new_zero_rate_P0  <- (par_value     / new_P0)^(1 / T) - 1
new_zero_rate_T   <- (par_value     / P0)^(1 / new_T) - 1

# Print new zero rates
cat("With a new par value of", new_par_value, ", the implied zero rate is:", round(new_zero_rate_par * 100, 2), "% p.a.\n")
cat("With a new price of", new_P0, ", the implied zero rate is:", round(new_zero_rate_P0 * 100, 2), "% p.a.\n")
cat("With a new term to maturity of", new_T, "years, the implied zero rate is:", round(new_zero_rate_T * 100, 2), "% p.a.\n")

```

### Bond Pricing Model II coupon bond 

Think of a coupon bond as a series of Zero-coupon bonds where the coupon is also a value like face value. 

$$
P_0 = \frac{C}{(1 + z_{0,1})} + \frac{C}{(1 + z_{0,2})^2} + \frac{C}{(1 + z_{0,3})^3} + \ldots + \frac{C}{(1 + z_{0,(T-1)})^{(T-1)}} + \frac{(C + \text{Par})}{(1 + z_{0,T})^T}
$$

where:

- $C$ is the coupon payment.
- $z_{0,t}$ is the zero rate for maturity $t$.
- $T$ is the total number of periods until maturity.
- $\text{Par}$ is the par value of the bond.

```{r}
# Define the parameters
coupons    <- c(100, 100, 100, 100, 1100)  # Coupon payments (including final par value)
zero_rates <- c(0.02, 0.025, 0.03, 0.035, 0.04)  # Corresponding zero rates
T <- length(coupons)  # Total number of periods

# Calculate the present value of each cash flow
present_values <- coupons / (1 + zero_rates)^(1:T)
present_values

# Sum the present values to get the bond price
bond_price <- sum(present_values)

# Print the bond price
cat("The price of the coupon bond is: $", round(bond_price, 2), "\n")
```

## Term Structure

### Yield Curve

The yield curve plots yields-to-maturity for coupon bonds against their terms to maturity. It's crucial to understand forward rates, zero rates, and actual rates.

## #Determinants of the Yield Curve Shape

| Factor                     | Description                                                                                      |
|----------------------------|--------------------------------------------------------------------------------------------------|
| Expectations Theory        | Influenced by expectations of future interest rates; upward if rates expected to rise.           |
| Liquidity Preference Theory| Longer-term bonds require a liquidity premium due to higher risk, leading to an upward slope.    |
| Market Segmentation Theory | Supply and demand in different maturity segments; can cause variations in yield curve shape.     |
| Inflation Expectations     | Higher expected inflation leads to higher yields on long-term bonds.                             |
| Central Bank Policies      | Short-term rates influenced by monetary policy; can steepen or flatten the yield curve.          |
| Economic Conditions        | Strong growth expectations lead to an upward slope; weak conditions can flatten or invert curve. |

### Reasons for Differing Yields by Maturity

| Factor                     | Description                                                                                      |
|----------------------------|--------------------------------------------------------------------------------------------------|
| Interest Rate Risk         | Longer-term bonds more sensitive to rate changes, requiring higher yields for compensation.      |
| Reinvestment Risk          | Short-term bonds face risk of reinvestment at lower rates; higher yields needed for longer terms.|
| Inflation Risk             | Longer maturities exposed to inflation risk, eroding value of future cash flows.                 |
| Credit Risk                | Longer-term bonds face more creditworthiness uncertainty; higher yields compensate for this risk. |
| Supply and Demand Dynamics | Investor preferences and liquidity needs affect demand and yields across maturities.             |

### Forward Rate Calculation
Forward rates can be derived from zero rates to determine arbitrage-free pricing.

$$
\begin{align}
f_{n,n+1} &= \left( \frac{(1 + z_{n+1})^{n+1}}{(1 + z_n)^n} \right) - 1
\end{align}
$$
To prevent arbitrage, we require that:

$$
X (1+z_{03})^3 \cdot (1+f_{35})^2 = X (1+z_{05})^5
$$

The general formula for the forward rate $f_{tT}$ (for $t < T$) is:

$$
f_{tT} = \left( \frac{(1+z_{0T})^T}{(1+z_{0t})^t} \right)^{\frac{1}{(T-t)}} - 1
$$

Where:

- $X$: Principal amount invested or lent.
- $z_{0t}$: Spot rate for a term ending at time $t$.
- $z_{0T}$: Spot rate for a term ending at time $T$.
- $f_{tT}$: Forward rate for the period between time $t$ and time $T$.
- $(1+z_{03})^3$: Accumulated value after 3 years with the spot rate $z_{03}$.
- $(1+z_{05})^5$: Accumulated value after 5 years with the spot rate $z_{05}$.
- $(1+f_{35})^2$: Accumulated value over 2 years with the forward rate $f_{35}$, starting at the end of year 3.

```{r}
# Define the spot rates and terms
z_05 <- 0.0725  # Spot rate for term t = 5 years
t    <- 5       # Term in years for spot rate z_05
z_08 <- 0.08755 # Spot rate for term T = 8 years
T    <- 8       # Term in years for spot rate z_08

# Calculate the forward rate f_{58}
f_58 <- ((1 + z_08)^T / (1 + z_05)^t)^(1 / (T - t)) - 1
f_58
```
### Term Structure of Interest Rates (TSIR)
TSIR describes the relationship between interest rates and terms. 

The three most often used interest rate curves are:

- $z$ : zero rate curves;
- $f$ : forward rate curves (for one-period investments); and
- $y$ : yield curves.


The yield curve can be influenced by:

1. **Expectations Hypothesis**: Investors' expectations about future interest rates.
2. **Liquidity Preference Theory**: Prefers short-term bonds for liquidity purposes.
3. **Market Segmentation Theory**: Different investors have different preferred investment horizons.
4. **Preferred Habitat Theory**: Investors have preferred investment horizons but will deviate for adequate compensation.

